Analysis of competitive surfing tournaments with generalized Bradley-Terry likelihoods

### 2021-08-26 inst/surfing_jsa.tex
more tweaks

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking; this would
suggest that more formal statistical methods are needed for analysis of
such datasets.  The Plackett-Luce likelihood functions are not subject
to such a phenomenon and furnish an objective and coherent method for
ranking competitors.  Further, likelihood-based methods offer one
advantage over points-based systems: the ability to conduct
statistically rigorous tests of meaningful nulls.

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\citep{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\citep{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  Such hypotheses are difficult to test with conventional
techniques as the competitors' strengths themselves are nuisance
parameters~\citep{basu1977}.

>>>>

Table~\ref{resultstable} shows the official results table from the
tour.  We see that Ferreira places first with 59740 points, Medina
second with 56475, and so on.  The overall ranking of the surfers is
determined by summing the season's points.  However, this system
suffers from the defect that the points system itself is intrinsically
arbitrary: the details of the points system does not affect the
competitors' behaviour but can change the overall ranking; this would
suggest that more formal statistical methods are needed for analysis
of such datasets~\citep{aldous2017}.  The Plackett-Luce likelihood
functions are not subject to such a phenomenon and furnish an
objective and coherent method for ranking competitors.  Further,
likelihood-based methods offer one advantage over points-based
systems: the ability to conduct statistically rigorous tests of
meaningful nulls.

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\citep{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\citep{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  In this context, risk sensitivity is likely to be become
important.  One example might be aerial manoeuvres which are risky
but, if successful, attract high scores~\citep{lundgren2014}.  Such
strategies are known to occur in chess~\citep{hankin2020}: players
suffering a sequence of defeats adopt a high-risk (all-or-nothing)
approach, while players in a winning streak hunker down, play
defensively, and entrench their lead.  Such hypotheses are difficult
to test with conventional techniques as the competitors' strengths
themselves are nuisance parameters~\citep{basu1977}.
